---
title: "Energy Market Volatility Analysis using MGARCH"
subtitle: "Project"
date: "`r Sys.Date()`"
author: 
  - name: "Evelina Vaitkevičiūtė <evelina.vaitkeviciute@mif.stud.lt>"
    title: DM (I) 1 k. - 2 variantas (No. 2525061)
output:
  readthedown::readthedown:
    highlight: kate
    fig_width: 14
    fig_height: 8
    number_sections: false
    code_folding: show
    logo: ./logo/MIF_EN.png
    css:
    - https://use.fontawesome.com/releases/v5.15.4/css/all.css
    - style.css
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(cache = TRUE)
```

This is a project for a Multivariate-Time-Series-and-Financial-Econometrics course, analysing the volatility dynamics and time-varying correlations between the stock and energy markets using a Multivariate GARCH (MGARCH) model. The analysis employs daily financial data obtained from Yahoo Finance through the `quantmod` package in R.

**Software and libraries**

```{r, echo=FALSE}
sessionInfo()
```

# Data

Below is given code To download the data set used in project.

```{r, error=FALSE, message=FALSE, warning=FALSE}
getData <- function(){
  
  library(quantmod)
  
  symbols_assets <- c("^GSPC", "BZ=F")
  
  for(sym in symbols_assets){
    
    x <- getSymbols(sym,
                    auto.assign = FALSE,
                    src = "yahoo",
                    from = "2016-01-01",
                    to   = "2026-04-01")
    
    clean_name <- gsub("\\^|=F", "", sym)
    
    assign(clean_name,
           na.omit(x),
           envir = .GlobalEnv)
  }
}
getData()
```

Two representative market indices are used:

- the **^GSPC** as a proxy for the United States stock market, and

```{r, echo=FALSE}
head(GSPC)
```

- the **BZ=F** as a proxy for the global energy market.

```{r, echo=FALSE}
head(BZ)
```

# Descriptive statistics

# Forecasting