---
title: "Energy and Stock Markets Volatility Analysis using MGARCH"
subtitle: "Project"
date: "`r Sys.Date()`"
author: 
  - name: "Evelina Vaitkevičiūtė <evelina.vaitkeviciute@mif.stud.lt>"
    title: DM (I) 1 k. - 2 variantas (No. 2525061)
output:
  readthedown::readthedown:
    highlight: kate
    fig_width: 14
    fig_height: 8
    number_sections: false
    code_folding: show
    logo: ./logo/MIF_EN.png
    css:
    - https://use.fontawesome.com/releases/v5.15.4/css/all.css
    - style.css
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(cache = TRUE)
```

This is a project for a Multivariate-Time-Series-and-Financial-Econometrics course, analysing the volatility dynamics and time-varying correlations between the stock and energy markets using a Multivariate GARCH (MGARCH) model. The analysis employs daily financial data obtained from Yahoo Finance through the `quantmod` package in R.

**Software and libraries**

```{r, echo=FALSE}
sessionInfo()
```

```{r, error=FALSE, message=FALSE, warning=FALSE}
library(quantmod)
library(rugarch)
library(rmgarch)
library(PerformanceAnalytics)
library(xts)
library(data.table)
library(ggplot2)
library(dplyr)
library(reshape2)
library(scales)
library(moments)
```

# Data

Below is given code to download the data set used in project.

```{r, error=FALSE, message=FALSE, warning=FALSE}
getData <- function(){
  
  library(quantmod)
  
  symbols_assets <- c("^GSPC", "BZ=F")
  
  for(sym in symbols_assets){
    
    x <- getSymbols(sym,
                    auto.assign = FALSE,
                    src = "yahoo",
                    from = "2016-01-01",
                    to   = "2026-04-01")
    
    clean_name <- gsub("\\^|=F", "", sym)
    
    assign(clean_name,
           na.omit(x),
           envir = .GlobalEnv)
  }
}
getData()
```

Two representative market indices are used:

- **S&P 500 index** (^GSPC) as a proxy for the United States stock market, and

```{r, echo=FALSE}
head(GSPC)
```

Dataset column names:
```{r}
ls(GSPC)
```

- **Brent Oil** (BZ=F) as a proxy for the global energy market.

```{r, echo=FALSE}
head(BZ)
```

Dataset column names:
```{r}
ls(BZ)
```

# Data preparation

*Consider that:*

- Time is consistent (i.e., ignore non-working days and periods with empty (`NA`) values),
- ignore exchange rate effects,
- $P_t^{(F)}$ - the closing price of the company's shares at time *t*

## Data filtering

First, need to extract Closing Prices:

```{r}
sp500 <- Cl(GSPC)
brent <- Cl(BZ)
```

Only those `GSPC` and `BZ=F` rows that do not have NA values are selected and are merged.

```{r}
data <- na.omit(merge(sp500, brent))
```

```{r, echo=FALSE}
head(data)
```

The final dataset consisted of **`r nrow(data)`** observations.

```{r}
data_old <- copy(data)
cols <- grep(".Close", colnames(data), value = TRUE)
data <- data.frame(time = index(data), value = data[, cols])
data <- as.data.table(data)

setnames(data, old = names(data), new = sub("^value\\.", "", names(data)))
```


```{r, echo=FALSE, fig.width=12, fig.height=5.5}

# Select closing prices
prices <- data[, c("time", "GSPC.Close", "BZ.F.Close")]

# Normalize prices (Base = 100)
prices$GSPC.Close <- prices$GSPC.Close / prices$GSPC.Close[1] * 100
prices$BZ.F.Close <- prices$BZ.F.Close / prices$BZ.F.Close[1] * 100

# Convert to long format
prices_long <- melt(
  prices,
  id.vars = "time",
  variable.name = "Market",
  value.name = "Price"
)

p <- ggplot(prices_long,
            aes(x = time,
                y = Price,
                color = Market)) +
  geom_line(linewidth = 1) +

  scale_color_manual(
    values = c(
      "GSPC.Close" = "navy",
      "BZ.F.Close" = "azure3"
    ),
    labels = c(
      "S&P 500",
      "Brent Oil"
    )
  ) +

  labs(
    x = "Time",
    y = "Index (Base = 100)",
    color = "Market"
  ) +

  theme_minimal() +

  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title.x = element_text(size = 17),
    axis.title.y = element_text(size = 17),
    axis.text.x = element_text(size = 16),
    axis.text.y = element_text(size = 16),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 15),
    legend.position = "top"
  )

print(p)

ggsave(
  "closing_prices.pdf",
  plot = p,
  width = 12,
  height = 5.5,
  dpi = 300
)
```

## Data transformation
Simple returns are calculated using the closing prices of company stocks and stock indices. Simple returns are expressed in logarithmic form and are further analyzed as log-returns.

```{r}
cols <- grep(".Close", colnames(data), value = TRUE)
data[, paste0(cols, ".simple") :=
      lapply(.SD, function(x) x / shift(x) - 1),
   .SDcols = cols]

data[, paste0(cols, ".log") :=
      lapply(.SD, function(x) log(x) - shift(log(x))),
   .SDcols = cols]
```

```{r}
data[, c("time", grep("simple|log", colnames(data), value = TRUE)), with = FALSE] %>% head(5)
```

The first value (row) is omitted because, when calculating the return on the very first day, we do not have a previous value.

```{r}
data <- data[-1, ]
```

```{r}
data[, c("time", grep("simple|log", colnames(data), value = TRUE)), with = FALSE] %>% head(5)
```




For further analysis will be used log returns:

```{r}
log_returns <- na.omit(diff(log(data_old)))
head(log_returns)
```

## Descriptive statistics
```{r}
desc_stats <- data.table(
  Variable = names(log_returns),
  
  Mean = sapply(log_returns, mean, na.rm = TRUE),
  Median = sapply(log_returns, median, na.rm = TRUE),
  Min = sapply(log_returns, min, na.rm = TRUE),
  Max = sapply(log_returns, max, na.rm = TRUE),
  Std_Dev = sapply(log_returns, sd, na.rm = TRUE),
  
  Skewness = sapply(log_returns, skewness, na.rm = TRUE),
  Kurtosis = sapply(log_returns, kurtosis, na.rm = TRUE)
)

print(desc_stats)

```

## Normality
```{r, echo=FALSE}
```{r fig.width=10, fig.height=5}
par(
  mfrow = c(1, 2),
  mar = c(4, 4, 2, 1)
)

sp500_returns <- as.numeric(log_returns$GSPC.Close)

h1 <- hist(
  sp500_returns,
  col = "#a5c8e0",
  border = "#104A7E",
  prob = TRUE,
  breaks = 50,
  xlab = "Log Returns",
  ylab = "Density",
  main = "",
  ylim = c(0, max(density(sp500_returns)$y) * 1.1)
)

d1 <- density(sp500_returns)
lines(d1$x, d1$y, col = "#868686", lwd = 2)
rug(sp500_returns, col = "#104A7E")
mtext("A", side = 3, adj = 0, line = 0.5, cex = 1.3, font = 2)

brent_returns <- as.numeric(log_returns$BZ.F.Close)

h2 <- hist(
  brent_returns,
  col = "grey80",
  border = "grey40",
  prob = TRUE,
  breaks = 50,
  xlab = "Log Returns",
  ylab = "Density",
  main = "",
  ylim = c(0, max(density(brent_returns)$y) * 1.1)
)

d2 <- density(brent_returns)
lines(d2$x, d2$y, col = "#868686", lwd = 2)
rug(brent_returns, col = "grey40")
mtext("B", side = 3, adj = 0, line = 0.5, cex = 1.3, font = 2)
```

Jarque-Bera Normality Test
```{r}
log_returns_SP500 <-  as.numeric(log_returns$GSPC.Close)
jarque.test(log_returns_SP500)
```
```{r}
log_returns_Brent <-  as.numeric(log_returns$BZ.F.Close)
jarque.test(log_returns_Brent)
```

Since p-value = $2.2 \times 10^{-16} < 0.05$, we reject the null hypothesis.

## Time series plots

```{r, echo=FALSE, error=FALSE, message=FALSE, warning=FALSE}
ggplot.corr <- function(data, lag.max = 24, ci = 0.95, large.sample.size = TRUE, horizontal = TRUE,...) {
  
  require(ggplot2)
  require(dplyr)
  require(cowplot)
  
  if(horizontal == TRUE) {numofrow <- 1} else {numofrow <- 2}
  
  list.acf <- acf(data, lag.max = lag.max, type = "correlation", plot = FALSE)
  N <- as.numeric(list.acf$n.used)
  df1 <- data.frame(lag = list.acf$lag, acf = list.acf$acf)
  df1$lag.acf <- dplyr::lag(df1$acf, default = 0)
  df1$lag.acf[2] <- 0
  df1$lag.acf.cumsum <- cumsum((df1$lag.acf)^2)
  df1$acfstd <- sqrt(1/N * (1 + 2 * df1$lag.acf.cumsum))
  df1$acfstd[1] <- 0
  df1 <- select(df1, lag, acf, acfstd)
  
  list.pacf <- acf(data, lag.max = lag.max, type = "partial", plot = FALSE)
  df2 <- data.frame(lag = list.pacf$lag,pacf = list.pacf$acf)
  df2$pacfstd <- sqrt(1/N)
  
  if(large.sample.size == TRUE) {
    plot.acf <- ggplot(data = df1, aes( x = lag, y = acf)) +
    geom_area(aes(x = lag, y = qnorm((1+ci)/2)*acfstd), fill = "#B9CFE7") +
    geom_area(aes(x = lag, y = -qnorm((1+ci)/2)*acfstd), fill = "#B9CFE7") +
    geom_col(fill = "#4373B6", width = 0.7) +
    scale_x_continuous(breaks = seq(0,max(df1$lag),6)) +
    scale_y_continuous(name = element_blank(), 
                       limits = c(min(df1$acf,df2$pacf),1)) +
    ggtitle("ACF") +
    theme_bw()
    
    plot.pacf <- ggplot(data = df2, aes(x = lag, y = pacf)) +
    geom_area(aes(x = lag, y = qnorm((1+ci)/2)*pacfstd), fill = "#B9CFE7") +
    geom_area(aes(x = lag, y = -qnorm((1+ci)/2)*pacfstd), fill = "#B9CFE7") +
    geom_col(fill = "#4373B6", width = 0.7) +
    scale_x_continuous(breaks = seq(0,max(df2$lag, na.rm = TRUE),6)) +
    scale_y_continuous(name = element_blank(),
                       limits = c(min(df1$acf,df2$pacf),1)) +
    ggtitle("PACF") +
    theme_bw()
  }
  else {
    plot.acf <- ggplot(data = df1, aes( x = lag, y = acf)) +
    geom_col(fill = "#4373B6", width = 0.7) +
    geom_hline(yintercept = qnorm((1+ci)/2)/sqrt(N), 
               colour = "sandybrown",
               linetype = "dashed") + 
    geom_hline(yintercept = - qnorm((1+ci)/2)/sqrt(N), 
               colour = "sandybrown",
               linetype = "dashed") + 
    scale_x_continuous(breaks = seq(0,max(df1$lag),6)) +
    scale_y_continuous(name = element_blank(), 
                       limits = c(min(df1$acf,df2$pacf),1)) +
    ggtitle("ACF") +
    theme_bw()
    
    plot.pacf <- ggplot(data = df2, aes(x = lag, y = pacf)) +
    geom_col(fill = "#4373B6", width = 0.7) +
    geom_hline(yintercept = qnorm((1+ci)/2)/sqrt(N), 
               colour = "sandybrown",
               linetype = "dashed") + 
    geom_hline(yintercept = - qnorm((1+ci)/2)/sqrt(N), 
               colour = "sandybrown",
               linetype = "dashed") + 
    scale_x_continuous(breaks = seq(0,max(df2$lag, na.rm = TRUE),6)) +
    scale_y_continuous(name = element_blank(),
                       limits = c(min(df1$acf,df2$pacf),1)) +
    ggtitle("PACF") +
    theme_bw()
  }
  cowplot::plot_grid(plot.acf, plot.pacf, nrow = numofrow)
}


```

```{r}
log_returns_gspc <- as.numeric(na.omit(diff(log(data_old[,"GSPC.Close"]))))
log_returns_brent <- as.numeric(na.omit(diff(log(data_old[,"BZ.F.Close"]))))
p <- ggplot.corr(
    data = log_returns_gspc,
    lag.max = 24,
    ci = 0.95,
    large.sample.size = TRUE,
    horizontal = TRUE
)
p
```

```{r, echo=FALSE}
ggsave(
    filename = "acf_pacf_panel.png",
    plot = p,
    width = 10,     # adjust to match your screenshot proportions
    height = 4,     # wider than tall
    dpi = 300
)
```

```{r}
Box.test(log_returns_gspc^2, lag = 10, type = "Ljung-Box")
Box.test(log_returns_brent^2, lag = 10, type = "Ljung-Box")
```

# MGARCH modelling

## Univariate GARCH(1,1)

```{r}
library(rugarch)

uspec <- ugarchspec(
  variance.model = list(model = "sGARCH", garchOrder = c(1,1)),
  mean.model = list(armaOrder = c(0,0)),
  distribution.model = "std"  # student-t for fat tails
)

fit_gspc  <- ugarchfit(uspec, data = log_returns_gspc)
fit_brent <- ugarchfit(uspec, data = log_returns_brent)

show(fit_gspc)
show(fit_brent)
fit_gspc
fit_brent
```

```{r}
# Standardized residuals
plot(fit_gspc,  which = 8)
plot(fit_brent, which = 8)

# Ljung-Box on residuals
Box.test(residuals(fit_gspc,  standardize = TRUE)^2, lag = 10, type = "Ljung-Box")
Box.test(residuals(fit_brent, standardize = TRUE)^2, lag = 10, type = "Ljung-Box")
```

## DCC-MGARCH
```{r}
library(rmgarch)

dcc_spec <- dccspec(
  uspec = multispec(replicate(2, uspec)),
  dccOrder = c(1,1),
  distribution = "mvt"
)

returns_matrix <- cbind(log_returns_gspc, log_returns_brent)
dcc_fit <- dccfit(dcc_spec, data = returns_matrix)
dcc_fit
```

```{r}
corr <- rcor(dcc_fit)

corr_series <- corr[1,2,]
plot(corr_series, type = "l",
     ylab = "Correlation", xlab = "Time")
```

# Forecasting
```{r}
# forecast horizon (e.g. 10 days)
dcc_forecast <- dccforecast(dcc_fit, n.ahead = 10)
rcor(dcc_forecast)
plot(dcc_fit, which = 4)
```
